Multifunctional Lipid Probes Background

### 2025-01-30 TrifunctionalLipidProbesOverview.qmd
removed naughty docx formatting

# Multifunctional Lipid Probes Background

@@

1.  Diazirine Rings – Enable covalent crosslinking to interacting proteins upon UV activation.
2.  Terminal Alkynes – Allow for affinity enrichment using click chemistry.
3.  Coumarin Photocages – Provide spatiotemporal control over lipid activity through light-triggered release.

#### Diazirine: Stabilizing Lipid-Protein Interactions with Covalent Crosslinking {#sec-DiazirinePlacement}

@@

#### Terminal Alkyne

@@

##### The Terminal Alkyne: A Versatile Bioorthogonal Handle

@@

By providing a simple yet effective means of modifying lipids without disrupting their native properties, the terminal alkyne has become an essential tool in lipidomics and chemical biology. It bridges the gap between traditional lipid research and modern bioorthogonal chemistry, allowing for precise studies of lipid function in living systems.

### Coumarin Photocage: Controlling Lipid Activity with Light

Lipids play an essential role in cellular signaling, yet their biological study presents unique challenges due to their highly dynamic metabolism. Unlike proteins, which are relatively stable once synthesized, many bioactive lipids are continuously synthesized, modified, interconverted, and degraded as part of an extensive and tightly regulated lipid metabolic network. This rapid turnover makes it difficult to study the effects of a specific lipid species in isolation, as the lipid of interest may be enzymatically altered before it can exert its intended biological function.

@@

#### A Light-Activated Protection Group

@@

#### Applications in Cellular Signaling Studies

@@

#### An Additional Benefit: Coumarin’s Inherent Fluorescence

@@

This dual functionality—serving as both a caging group and a fluorescent probe—makes coumarin a useful tool for studying lipid localization, uptake, and bioactivity.

@@

#### Label-Free Quantification

@@

Label-free quantification is a practical and widely used method for proteomics research. While cost-effective and flexible, its accuracy depends on careful experimental design and quality control to reduce variability.

#### Quantitative Mass Spectrometry with Tandem Mass Tagging {#sec-TMT}

@@

TMT-based mass spectrometry is a valuable tool for large-scale proteomics, offering high-throughput, precise, and deep proteome coverage. While it requires advanced instrumentation and careful sample preparation, its ability to multiplex multiple samples in a single run makes it an attractive option for high-confidence protein quantification.

>>>>

@@

:::{.callout-note collapse="true"}
#### Diazirine ring

@@

##### The effect of Diazirine placement on crosslinking targets {#sec-DiazirinePlacement}

- The reactive carbene produced upon irradiation will bond with whatever molecule is nearest -- thus, the location of the diazirine ring can have a dramatic effect on what protein interactors are stabilized via crosslinking.

- The majority of lipid analogs produced to date were synthesized with the diazirine ring near the terminal end of an acyl chain. For example, [8-3 Fatty Acid](/LipidProbe/8-3_FattyAcid.qmd) has the acyl chain only 3 carbons from the end. These lipid analogs may be expected to preferentially crosslink to proteins which penetrate far into the hydrophobic region between the leaflets of a membrane bilayer, such as transmembrane proteins.

- In comparison, moving the diazirine ring closer to the hydrophilic headgroup may increase the partiality of crosslinking towards peripheral membrane proteins in addition to transmembrane proteins. An example of this shift is [1-10 Fatty Acid](/LipidProbe/1-10_FattyAcid.qmd), in which the diazirine ring was placed 10 carbons from the terminus of the chain.
:::

:::{.callout-note collapse="true"}
#### Terminal Alkyne

@@

##### A Versatile Bioorthogonal Handle

@@

By providing a simple yet effective means of modifying lipids without disrupting their native properties, the terminal alkyne has become an essential tool in lipidomics and chemical biology. It bridges the gap between traditional lipid research and modern bioorthogonal chemistry, allowing for precise studies of lipid function in living systems.
:::

:::{.callout-note collapse="true"}
#### Coumarin Photocage
Lipids play an essential role in cellular signaling, yet their biological study presents unique challenges due to their highly dynamic metabolism. Unlike proteins, which are relatively stable once synthesized, many bioactive lipids are continuously synthesized, modified, interconverted, and degraded as part of an extensive and tightly regulated lipid metabolic network. This rapid turnover makes it difficult to study the effects of a specific lipid species in isolation, as the lipid of interest may be enzymatically altered before it can exert its intended biological function.

@@

##### A Light-Activated Protection Group

@@

##### Applications in Cellular Signaling Studies

@@

##### An Additional Benefit: Coumarin’s Inherent Fluorescence

@@

This dual functionality—serving as both a caging group and a fluorescent probe—makes coumarin a useful tool for studying lipid localization, uptake, and bioactivity.
:::

@@

:::{.callout-note collapse="true"}
#### Label-Free Quantification

@@

Label-free quantification is a practical and widely used method for proteomics research. While cost-effective and flexible, its accuracy depends on careful experimental design and quality control to reduce variability.
:::
:::{.callout-note collapse="true"}
#### Quantitative Mass Spectrometry with Tandem Mass Tagging {#sec-TMT}

@@

TMT-based mass spectrometry is a valuable tool for large-scale proteomics, offering high-throughput, precise, and deep proteome coverage. While it requires advanced instrumentation and careful sample preparation, its ability to multiplex multiple samples in a single run makes it an attractive option for high-confidence protein quantification.
:::
